Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-19 draft.tex
.

\subsubsection{Distribution of Synaptic Weights}

@@

In contrast to our model, Effenberger et. al. identified the presence of inhibitory STDP as a necessary mechanism for the development of this feature. On a deeper explanatory level, the authors argue that inhibitory STDP is indirectly involved: it weakens inhibitory currents for some neurons, which cause these neurons to have more above-average firing rates. In turn, synapses between highly active presynaptic cells and postsynaptic neurons with lower activity are known to be subject to long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}. With these causal relations in mind, we can argue that diffusive homeostasis effectively embodies the same functional role as inhibitory STDP in \cite{Effenberger_2015} by allowing for the presence of excitatory cells with above-average activity. We tested this relationship by plotting mean outgoing weights against the average firing rate, see Figure \ref{Out_Weight_vs_F}. A strong heterogeneity of firing rates allows for the development of few strong synapses, while the resulting distribution of mean weights of a narrow distribution is limited to a smaller range.
\begin{figure}
%%% out_weight_vs_f.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/mean_out_weights_vs_rate.png}
\caption{Mean of outgoing excitatory weights at $\mathrm{t=1500\,s}$ plotted against mean presynaptic firing rate (averaged over $\mathrm{t=1200-1500\,s}$) in regular (A) and logarithmic space (B). Each plot corresponds to a presynaptic excitatory cell.}
\label{Out_Weight_vs_F}
\end{figure}
It is interesting to note that on a logarithmic weight scale, we found a roughly linear relationship. Figure \ref{Out_Weight_vs_F} only depicts data from simulations including a distance dependent connection profile, since removing this topology did not result in noticeable differences.

@@

\subsection{Equilibrium in the full Diffusive Model}
Since the previous section has shown that the simplified model in \cite{Sweeney_Paper} has problems \emph{maintaining} a broad distribution of firing rates, turning to a more general formulation of the problem seemed reasonable.

@@

\subsubsection{Possible Modifications}
Since these considerations evoked a rather sceptical view onto the biological plausibility of these results, one should consider possible modifications of the homeostatic model. One should note that these are to be evaluated in the context of providing a model of an actual physiological process. If the main goal was to simply provide a homeostatic method that retains an overall level of activity without tightly fixing individual rates, we would suggest to use the idealized model of instant diffusion, especially because of its computational simplicity. However, as already mentioned, the given experimentally based parameters do not allow to accept this simplification from the standpoint of biological and physical accuracy. Two elements of the model should be put into question: First, is it appropriate to assume an approximately point-like source of nitric oxide? Second, are the dynamics of threshold adaption (Equation \eqref{Theta_dyn}) an acceptable description of the actual physiological processes?

@@

A major issue of Equation \eqref{Theta_dyn} is the fact that, given a clamped positive or negative difference between the actual NO concentration and the target concentration, the excitability of the neuron would decrease or increase without bounds. Because of the limited amount of ion channels that exist in a cell and can be affected by nitric oxide, upper and lower bounds should certainly exist. A simple approach to tackle this would be to implement hard bounds into our neuron model, not allowing the threshold to pass an upper and lower value. Another possibility could be to implement these bounds by means of the differential equation itself.

In either case, such a modification would break the ability of the network to tune thresholds in such a way that \emph{all} neurons eventually reach target concentration.

>>>>

\subsubsection{Distribution of Synaptic Weights} \label{Syn_Weight_Dist_Section}

@@

In contrast to our model, Effenberger et. al. identified the presence of inhibitory STDP as a necessary mechanism for the development of this feature. On a deeper explanatory level, the authors argue that inhibitory STDP is indirectly involved: it weakens inhibitory currents for some neurons, which cause these neurons to have more above-average firing rates. In turn, synapses between highly active presynaptic cells and postsynaptic neurons with lower activity are known to be subject to long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}. With these causal relations in mind, we can argue that diffusive homeostasis effectively embodies the same functional role as inhibitory STDP in \cite{Effenberger_2015} by allowing for the presence of excitatory cells with above-average activity. We tested this relationship by plotting mean outgoing weights against the average firing rate, see Figure \ref{Out_Weight_vs_F}. A strong heterogeneity of firing rates allows for the development of few strong synapses, while the resulting distribution of mean weights of a narrow distribution is limited to a smaller range. We also tested the case of instant diffusion. Though firing rates span across a similar range, the resulting mean weights do not reach as high values as in the case of diffusion at a finite rate.
\begin{figure}
%%% out_weight_vs_f.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/mean_out_weights_vs_rate.png}
\caption{Mean of outgoing excitatory weights at $\mathrm{t=1500\,s}$ plotted against mean presynaptic firing rate (averaged over $\mathrm{t=700-1000\,s}$) in regular (A) and logarithmic space (B). Each plot corresponds to a presynaptic excitatory cell.}
\label{Out_Weight_vs_F}
\end{figure}
Furthermore, it is interesting to note that on a logarithmic weight scale, we found a roughly linear relationship. Figure \ref{Out_Weight_vs_F} only depicts data from simulations including a distance dependent connection profile, since removing this topology did not result in noticeable differences.

@@

\subsection{Equilibrium in the full Diffusive Model} \label{Matrix_Diff_Model_Section}
Since the previous section has shown that the simplified model in \cite{Sweeney_Paper} has problems \emph{maintaining} a broad distribution of firing rates, turning to a more general formulation of the problem seemed reasonable.

@@

\subsubsection{Possible Modifications}
Since these considerations evoked a rather sceptical view onto the biological plausibility of these results, one should consider possible modifications of the homeostatic model. One should note that these are to be evaluated in the context of providing a model of an actual physiological process. If the main goal was to simply provide a homeostatic method that retains an overall level of activity without tightly fixing individual rates, we would suggest to use the idealized model of instant diffusion, especially because of its computational simplicity. However, as already mentioned, the given experimentally based parameters do not allow to accept this simplification from the standpoint of biological and physical accuracy. Instead, two elements of the model could be put into question: First, is it appropriate to assume an approximately point-like source of nitric oxide? Second, are the dynamics of threshold adaption (Equation \eqref{Theta_dyn}) an acceptable description of the actual physiological processes?

@@

A major issue of Equation \eqref{Theta_dyn} is the fact that, given a clamped positive or negative difference between the actual NO concentration and the target concentration, the excitability of the neuron would decrease or increase without bounds. Because of the limited amount of ion channels that exist in a cell and that can be affected by nitric oxide, upper and lower bounds should certainly exist. A simple approach to tackle this would be to implement hard bounds into our neuron model, not allowing the threshold to pass an upper and lower value. Another possibility is the implementation of these bounds by means of the differential equation itself.

In either case, such a modification would break the ability of the network to tune thresholds in such a way that \emph{all} neurons eventually reach target concentration. As a consequence, the fixed point of firing rates under diffusive homeostasis would not be exclusively determined by spatial structure, but by a combination of network structure and the neurons' positions. As an example, consider a modification of Equation \eqref{Theta_dyn} of the form
\begin{equation}
\dot{V_t^i}(t) = \frac{1}{\tau_{V_t}}\left(\frac{NO(\mathbf{r}_{neur}^i,t)-NO_0}{NO_0} + \frac{\alpha}{V_t^i - V_{t,min}} + \frac{\alpha}{V_t^i - V_{t,max}}\right) \; .
\label{Theta_dyn_with_Bounds}
\end{equation}
In this form, $\mathrm{NO(\mathbf{r}_{neur}^i,t) = NO_0}$ does not follow any more as the fixed point solution. Rather, one needs to consider the fact that, given the current network topology, the steady state of $\mathrm{NO(\mathbf{r}_{neur}^i)^* = NO(\mathbf{r}_{neur}^i,V_t^1,...,V_t^n)^*}$ in turn depends on the set of thresholds. Unfortunately, this increase in complexity would presumably make it impossible to find a general analytic description of the fixed point as introduced and discussed in Section \ref{Matrix_Diff_Model_Section}. Moreover, while this modification would incorporate network topology into the steady state of the system for finite diffusion constants, it is not clear whether it would allow perturbations (e.g. due to external input) to be sustained longer.

\subsection{Network Topology} 
Even though our investigations on network topology were rather intended as a form of sanity-check with respect to properties already found in earlier versions, we found some differences compared to non-diffusive homeostasis.

First, in Section \ref{Syn_Weight_Dist_Section}, we presented our results regarding the distribution of weights, which turned out to be broader but still approximately log-normal distributed for diffusive homeostasis. As a first interpretation, we hypothesized that this might due to a slightly decreased amount of STDP-induced potentiation (Figure \ref{STDP_change}) in the case of diffusive homeostasis in the first half of the simulation. Another explanation one might consider is that the increased firing rate for some neurons in the case of diffusive homeostasis leads to a greater amount of fluctuation because of the increased amount of STDP-events. However, since the overall variance of STDP-induced fluctuations is not significantly higher but even slightly lower compared to non-diffusive homeostasis, this explanation seems rather unlikely.

Taking a closer look at the distributions depicted in Figure \ref{Weight_Dist}, we noted that
